---
title: "INF 511 Assignment 3"
author: "Jen Diehl (6179236), Sam Watson (6174574), Natasha Wesely (6180693)"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
options(knitr.kable.NaN = '')
```


```{r Set Up}
hw3.df<- readRDS(file="hw3.df.RDS")
plot(log(H/PD) ~ x, data=hw3.df, pch=as.numeric(HVI),
      xlab=expression("Daily Maximum Temperature" ~ (x[i])),
      ylab=expression("Hospitalization Rate" ~ (log(H[i]/PD[i]))))
legend("topleft", legend=c("high", "med", "low"), pch=c(3,2,1),title="HVI Group")
```

## Question 1

```{r Q1}
lmod <- lm(log(H/PD)~I(x-30)+I((x-30)^2)+HVI ,data = hw3.df)
summary(lmod)
```

## Question 2 
```{r Q2}
summary(lmod)$r.squared
```
The proportion of variability of log hospitalization rate is 0.8995752 explained by the association of temperature and HVI.

## Question 3
The intercept from this model output is around -16. If we were to draw a line for the data on the plot above the intercept is also around -16. In the model we have shifted the y-axis to the right where 30 now equals 0 (or where the 20 is on the above plot). Therefore, it is appropriate to infer about the intercept from the model.

## Question 4
### i
H$_0$: y = $\beta_{0}$ + $\beta_{1}$ * (Temperature - 30) + $\beta_{2}$ * (Temperature - 30)$^2$

H$_a$: y = $\beta_{0}$ + $\beta_{1}$ * (Temperature - 30) + $\beta_{2}$ * (Temperature - 30)$^2$ + $\beta_{3}$  * HVIM + $\beta_{4}$ * HVIH

```{r Q4}
lmod2 <- lm(log(H/PD)~I(x-30)+I((x-30)^2) ,data = hw3.df)
summary(lmod2)
modanova <- anova(lmod, lmod2)
anova(lmod, lmod2)
```
### ii
The F statistic is 24.847.

### iii
The p-value is 1.487e-08.

### iv
We can reject our null hypothesis that the reduced model is sufficient because our p-value is less than 0.05.

## Question 5
### i
There are 5 parameters in the full model.

### ii
There are 3 parameters in the reduced model.

### iii
For the full model DF is 59 and the reduced model DF is 61.

### iv

```{r Q5 iv}
pval = 1 - pf(modanova$F[2],
              df1 = modanova$Res.Df[1], 
              df2 = modanova$Res.Df[2])
print(pval)
```
The p-value is very small number and R is rounding to zero.

## Question 6

```{r Q6}
b4hat = summary(lmod)$coefficients[5,1]
seB4hat = summary(lmod)$coefficients[5,2]
tcrit = qt(0.95, df = 59)

(CIlower = b4hat - (tcrit*seB4hat))
```

We are 95% confident the mean effect of high HVI (relative to low HVI) is at least 0.8692906 (or greater).  

## Question 7
H$_0$: $\beta_{4}$ = 0

H$_a$: $\beta_{4}$ > 0 

```{r}
b4hat = summary(lmod)$coefficients[5,1]
seB4hat = summary(lmod)$coefficients[5,2]

testStat = (b4hat - 0) / seB4hat
(pval = 1 - pt(testStat, df = 59))
```

The test statistic is far into the right tail of the t distribution (test stat t-value = 7.027036), and p-value is close to zero (1.217393e-09). Therefore, we can reject the null hypothesis (that there is no effect of high HVI). The effect of high HVI is significantly greater than zero. 
