---
title: "INF 511 Assignment 2"
author: "Jen Diehl (6179236), Sam Watson (6174574), Natasha Wesely (6180693)"
date: "2022-09-16"
output: pdf_document
---


```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
options(knitr.kable.NaN = '')
```

## Question 1
### (a)

```{r Question 1a}
sapflow.df<-  readRDS(file="sapflow.rds")
str(sapflow.df)
```
V2 seems to be a binary variable that would most likely take on a binomial distribution because there are only two options.

### (b)
```{r Question 1b}
names(sapflow.df) = c("light","fertilizer","temperature","moisture","sapflow")
head(sapflow.df)
```

### (c)
```{r Question 1c}
sapflow.df$light = factor(sapflow.df$light, labels = c("lo","hi"))
head(sapflow.df)
str(sapflow.df)
```

### (d)
```{r Question 1d}
pairs(sapflow.df)
```

The pairs function reports the factors as their numeric value of 1 or 2.

### (e)
A single histogram may not be enough to assess if sapflow is normally distributed because it is just a visual assessment of normality which is inherently subjective. Creating a histogram is not formally or statistically assessing normality.

## Question 2

### (a)
```{r Question 2a}
X<- model.matrix(sapflow ~ light + fertilizer + temperature + moisture, data=sapflow.df)
y<- sapflow.df$sapflow
```

E(Y$_1$ | x$_1$) = ${\beta}_0$ * X$_{1,0}$ + ${\beta}_1$ * X$_{1,1}$ + ${\beta}_2$ * X$_{1,2}$ + ${\beta}_3$ * X$_{1,3}$ + ${\beta}_4$ * X$_{1,4}$

E(Y$_1$ | x$_1$) = ${\beta}_0$ * 1 + ${\beta}_1$ * 0 + ${\beta}_2$ * 1 + ${\beta}_3$ * 0 + ${\beta}_4$ * 1

E(Y$_1$ | x$_1$) = ${\beta}_0$ + ${\beta}_2$ + ${\beta}_4$


### (b)

E(Y$_5$ | x$_5$) = ${\beta}_0$ * X$_{5,0}$ + ${\beta}_1$ * X$_{5,1}$ + ${\beta}_2$ * X$_{5,2}$ + ${\beta}_3$ * X$_{5,3}$ + ${\beta}_4$ * X$_{5,4}$

E(Y$_5$ | x$_5$) = ${\beta}_0$ * 1 + ${\beta}_1$ * 0 + ${\beta}_2$ * 1 + ${\beta}_3$ * 1 + ${\beta}_4$ * 1

E(Y$_5$ | x$_5$) = ${\beta}_0$ + ${\beta}_2$ + ${\beta}_3$ + ${\beta}_4$

E (Y$_5$ | x$_5$) - E(Y$_1$ | x$_1$) = ${\beta}_0$ + ${\beta}_2$ + ${\beta}_3$ + ${\beta}_4$ - ${\beta}_0$ - ${\beta}_2$ - ${\beta}_4$

E (Y$_5$ | x$_5$) - E(Y$_1$ | x$_1$)  = ${\beta}_3$

The difference mean sapflow between high and low temperature and other inputs is equal to ${\beta}_3$.

### (c)
```{r Question 2c}
model = lm(sapflow ~ light + fertilizer + temperature + moisture, data = sapflow.df)
summary(model)
```
The model tells us that when moving from low to high temperature there is a 43.92 mL/h increase in sapflow.

### (d)
#### i.
$\hat{\beta}_3$ = 43.92

#### ii.
SE $\hat{\beta}_3$ = 13.17

#### iii.
t-value $\hat{\beta}_3$ = 3.334

#### iv.
11 degrees of freedom

#### v.
p-value = 0.006660

#### vi.
Yes, the effect of temperature is statistically significant because the t-value is far into the tails of the t distribution, therefore the p-value is small and less than alpha. 

### (e)
```{r Question 2e}
xtxi = solve(t(X) %*% X)
betahat = xtxi %*% t(X) %*% y
betahat
coef(model)
```
The beta hats are very similar from the manual calculations and the model created with lm.

### (f)
```{r Question 2f}
S = t(X)%*%X
Xinv = solve(S)
H = X %*% Xinv %*% t(X)
H
```

### (g)
```{r Question 2g}
yhat = H%*%y
yhat
fitted(model)
all(round(yhat - fitted(model), 10)==0)
```
Yes, the yhats calculated from the matrix-vector computations match the yhats from the model created with lm().

### (h)
```{r Question 2h}
resids = y-yhat
all(round(resids - residuals(model), 10)==0)
```
Yes, the residuals from the matrix-vector computation match the residuals from the model created with lm().

### (i)
```{r Question 2i}
sd(resids)
```
From the model the residual standard error: 26.35 

The standard deviation of errors from our matrix-vector computation is slightly different from the model made with lm(). The difference in the estimates is reasonable considering there are only 16 observations.

### (j)
```{r Question 2j}
xtxi <- solve(t(X) %*% X)
varHat = sum(((y-yhat)^2) / 11 )
varB = xtxi * varHat
all(round(varB - vcov(model), 10)==0)
```

Yes, our matrix computation of $Var(\hat{\beta})$ is the same as vcov(model).

```{r}
sigHat = sqrt(varHat)
(seHatB3 = sqrt(diag(xtxi)) * sigHat)[4]

summary(model)$coefficients[4,2]
```

Yes, the matrix computation of the $\hat{se}(\hat{\beta_{3}})$ matches the $\hat{se}(\hat{\beta_{3}})$ from the lm() object.

### (k)
```{r Question 2k}
(R2 = cor(y, yhat)^2)
summary(model)$r.squared
```

Yes, our computed value of R2 agrees with the output from our previously computed lm() model.
